Higher-Order unification for free!

### 2024-04-12 paper/main.tex
typos

Even if we encountered the problem working on CIC we devise
a minimal setting to ease its study. In this setting we have
a \Fo{} language (for first order) with a richer equational
theory and a \Ho{} meta language with a simpler one.

@@

\noindent
When it is clear from the context we shall use the same syntax for \Fo{} terms
(although we never subscripts unification variables).

We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
terms in \Ho{}.

@@

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app[F|L]}>>.

@@

Although we provide an implementation of the meta-language unification \Ue
in the supplementary material (that we used for testing purposes) we only
describe its signature here.

@@

We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$.
We write $t_1 \Ue t_2 \mapsto \sigma'$ when
the initial substitution $\sigma$ is empty.
% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
% the domains of $\sigma$ and $\sigma'$ are disjoint.
We write \llambda as the set of terms that are in the pattern-fragment, i.e.
every higher-order variable is applied to a list of distinct names.

@@

Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
logic program.

\subsection{The problem: logic-program simulation}
% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
\foUnifPb is a list of pair of terms, each pair correspond to a unification
problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
elements extracted from $P_p$).
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

>>>>

Even if we encountered the problem working on CIC we devise
a minimal setting to ease its study. In this setting we have
a \Fo{} language (for first order) with a rich equational
theory and a \Ho{} meta language with a simpler one.

@@

\noindent
When it is clear from the context we shall use the same syntax for \Fo{} terms
(although we never subscripts unification variables).\\
We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
terms in \Ho{}.

@@

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app [F|L]}>>.
Note that the symbol \elpiIn{|} separates the head of a list from the tail.

@@

% Although we provide an implementation of the meta-language unification \Ue
% in the supplementary material (that we used for testing purposes) we only
% describe its signature here.
\Ho{}'s unification signature is:

@@

\noindent
We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
reflected by signature above that relates two substitutions.
We write $t_1 \Ue t_2 \mapsto \sigma'$ when
the initial substitution $\sigma$ is empty.
% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
% the domains of $\sigma$ and $\sigma'$ are disjoint.
We write \llambda as the set of terms that are in the pattern-fragment, i.e.
every higher-order variable is applied to a list of distinct names.

@@

\noindent
Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
logic program.

\subsection{The problem: logic-program simulation}
% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
\foUnifPb is a list of pair of terms, each pair corresponds to a unification
problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
elements extracted from $\foUnifPb_p$).
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$
